---
title: "Causes, inferences, and peril"
author: "Austin Hart"
output: 
  ioslides_presentation:
    widescreen: true
    css: iocustom.css
---

```{r setup, include=FALSE}
  options(htmltools.dir.version = FALSE)
  knitr::opts_chunk$set(
    message=F, warning=F, eval=T, echo = F, 
    fig.align = 'center'
  )

  library(tidyverse)  
  library(knitr)
```


# Thinking scientfically

## The process

- Causal questions
  - **Theory** as baseline
  - **Hypotheses** as specific predictions
  
- Observation
  - Research design
  - Data collection
  
- Analysis
  
- Inference (not answers, truth, etc.)
  
  
## Inferences and validity

> What is the effect of exposure to X on outcome Y?  
> Based on my study, I can say...

- **Statistical inference:** formal judgment about the parameters of a population; typically for differentiating coincidental from systematic correlations.  

- **Causal inference (internal):** attribution of an effect to a cause based on evidence *from the study*.

- **Generalization (external):** inference about how well the observed effect extends to other places, times, treatments, outcomes, and/or units *beyond the study*.


## Consider two examples

- Cohen and Dupas (2010)
- Keele et al. (2021)

For each study:
- What is the motivating question/theory?
- What effect are they trying to estimate?
- Describe the study design
- Discuss the internal/external validity of their findings


# Fundamental Problem of Causal Inference

## Define a causal effect

For an individual, $i$, under study, we could observe:
- $Y_{i,A}$: $i$'s response to $Treat = A$
- $Y_{i,B}$: $i$'s response to $Treat = B$

Calculate the causal effect of $Treat$ on $Y$ as: 
$$
 = Y_{i,A} - Y_{i,B}
$$

## Thinking carefully about treatment effects

$$
 = Y_{i,A} - Y_{i,B}
$$

- "effect of treatment" is always comparative
  - "The effect of $A$ (compared to $B$) on Y"
  - The referent changes everything

- "cause" requires a counterfactual
  - The unseen potential outcome
  - What *would have happened if* the unit were exposed instead to $T=?$
  

## Why causality is unobservable

> We can observe a unit's response to treatment $(Y_{i,A})$. 
> Or... we can observe the response of an untreated unit $(Y_{i,B})$. 
> We cannot observe both for the same unit at the same time.  

- We can record, at most, one potential outcome.
- **IMPOSSIBLE to observe a causal effect directly**
- Must rely on causal INFERENCE


## ## Approximating a counterfactual?  

### Scientific solutions
1. Compare response across two units, $i$ and $j$: 
$$Y_{i,A} - Y_{j,B}$$
2. Expose unit to both in sequence: 
$$Y_{i,A,t} - Y_{i,B,t+1}$$


> What assumptions do we need for these comparisons to be valid?


## Approximating a counterfactual? 

### Statistical solution
- Estimate the **Average Treatment Effect** over a large number of units.
  - Average outcome among the treated, $\bar{Y_A}$ 
  - Average outcome among untreated, $\bar{Y_B}$

$$ATE = \bar{Y_A}-\bar{Y_B}$$

> What assumptions must we make for this comparison to be valid?


# FPCI in practice

## Ruling out alternative explanations

> I claim that, across nations, wealth (X) increases life expectancy (Y). Is this a valid causal claim?

### Criteria for internal validity
- Demonstrate correlation between $X$ and $Y$
- Rule out random coincidence: causal mechanism
- Rule out bias due to reverse causality $(Y \rightarrow X)$
- Rule out bias due to confounding $(X \leftarrow Z \rightarrow Y)$


## Correlation

> Correlation tells us about the tendency of two features of the world (variables) to occur together or not.

- **Positive correlation**: two features of the world (or variables) tend to occur together. 

- **Negative correlation**: when one feature occurs, the other tends not to occur.

- **Independence (no correlation)**: the occurrence of one feature is unrelated to the presence or absence of the other.


## Example 1

```{r gapplot, fig.dim=c(5,3)}
gapminder::gapminder %>%
  filter(year == 1997) %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point(shape = 21, color = 'red') +
  geom_smooth(method = 'loess', se = F, color = 'black') +
  labs(x = 'GDP per capita',
       y = 'Life expectancy, years',
       title = 'Wealth drives life expectancy',
       caption = 'Data: Gapminder, 1997') +
  scale_x_log10(labels = scales::label_dollar()) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())

```

## Example 2

```{r gapplot2, fig.dim=c(5,3)}
gapminder::gapminder %>%
  filter(year == 1997) %>%
  ggplot(aes(x = pop, y = lifeExp)) +
  geom_point(shape = 21, color = 'red') +
  geom_smooth(method = 'loess', se = F, color = 'black') +
  labs(x = 'Population',
       y = 'Life expectancy, years',
       title = 'Life expectancy by population',
       caption = 'Data: Gapminder, 1997') +
  scale_x_log10(labels = scales::label_log()) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())

```

## Bias due to reverse causality

- Exists when the outcome is acting on the proposed cause. $X \leftrightarrow Y$
- If $Y$ determines exposure to $X$, then the correlation reflects the effect of $Y$ on $X$ in full or in part.

Could life expectancy be acting on GDP?
```{r reverse, fig.dim=c(5,3)}
gapminder::gapminder %>%
  filter(year == 1997) %>%
  ggplot(aes(y = gdpPercap, x = lifeExp)) +
  geom_point(shape = 21, color = 'red') +
  geom_smooth(method = 'loess', se = F, color = 'black') +
  labs(y = 'GDP per capita, $',
       x = 'Life expectancy, years',
       title = 'Life expectancy drives wealth?',
       caption = 'Data: Gapminder, 1997') +
  scale_y_log10(labels = scales::label_dollar()) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())

```


## Bias due to confounding

> If X and Y are correlated, could it be the result of a third variable, Z, acting on both?


- Selection/confounding bias exists when:
  - Treatment groups differ systematically by some factor, $Z$
  - *AND* $Z$ is a cause of $Y$
- In this case, the estimated effect of $X$ on $Y$ is biased due to the confounding effect of $Z$.


## Back to the readings

1. How do the designs attempt to maximize internal validity?
2. What is the effect of random assignment in the Cohen and Dupas study?

