---
title: "Inferences"
subtitle: "and how to assess them"
author: "Austin Hart"
institute: "American University"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, rladies, rladies-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  message=FALSE, warning=FALSE, eval = TRUE, echo = FALSE, 
  fig.align = 'center', dev='svglite', fig.path = "images/"
)
  
  library(tidyverse)
```

class: inverse, middle, left

# Thinking scientfically  


---
# The process  

- Causal questions

- Theory and hypotheses
  - **Theory** as baseline
  - **Hypotheses** as specific predictions
  
- Data gathering
  - Research design
  - Data collection
  
- Analysis
  
- Inference 
  
  
  
---
# Inferences and validity

> I gathered data for a study. For the cases in my sample, I find a relationship between two variables, X and Y.

- **Causal inference:** attribution of an effect to a cause *within a study*. I infer that exposing the units in my sample to X caused a change in Y.

- **External validity inference:** extending the results *beyond the study* to other:  
  - populations  
  - treatments  
  - outcomes  
  - contexts  


---
class: inverse, middle, left

# Fundamental Problem of Causal Inference


---
# Define a causal effect

For an individual, $i$, under study, we could observe 

- $Y_{i,T=1}$, $i$'s response after treatment $(T=1)$
- $Y_{i,T=0}$, $i$'s response after control $(T=0)$

Calculate the causal effect of treatment on outcome $Y$ as:  

$$Y_{i,T=1} - Y_{i,T=0}$$


---
# Why causality is unobservable

> We can observe a unit's response to treatment  $(Y_{i,T=1})$. 
> Or... we can observe the response without treatment $(Y_{i,T=0})$. 
> We cannot observe both for the same unit at the same time.  

- We can observe, at most, one of two potential outcomes.  

- The unobserved outcome is called the counterfactual.  

- **IMPOSSIBLE to observe a causal effect directly**


---
# Recovering the counterfactual

### Scientific solutions  

- Compare response across two units, $i$ and $j$: $Y_{i,T=1} - Y_{j,T=0}$  
- Expose unit to both treatments in sequence: $Y_{i,T=1,t} - Y_{i,T=0,t+1}$    


### Statistical solution

- Estimate the **Average Treatment Effect** over a large number of sampled units.

$$SATE = Avg_n[Y_i|T_i=1] - Avg_n[Y_i|T_i=0]$$


---
class: inverse, middle, left

# FPCI in practice


---
# Establishing internal validity 

### Conclude that $X$ is causing $Y$ if you can:  

- Specify a plausible causal mechanism  

- Demonstrate correlation between $X$ and $Y$   

- Rule out bias due to reverse causality   

- Rule out bias due to confounding   



---
# Correlation

> Correlation tells us about the tendency of two features of the world (variables) to occur together.

- **Positive correlation**: two features of the world (or variables) tend to occur together. 

- **Negative correlation**: when one feature occurs, the other tends not to occur.

- **Independence (no correlation)**: the occurrence of one feature is unrelated to the presence or absence of the other.

---
# Example 1

```{r gapplot, fig.dim=c(6,4)}
gapminder::gapminder %>%
  filter(year == 1997) %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point(shape = 21, color = 'red') +
  geom_smooth(method = 'loess', se = F, color = 'black') +
  labs(x = 'GDP per capita',
       y = 'Life expectancy, years',
       title = 'Wealth drives life expectancy',
       caption = 'Data: Gapminder, 1997') +
  scale_x_log10(labels = scales::label_dollar()) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())

```

---
# Example 2

```{r gapplot2, fig.dim=c(6,4)}
gapminder::gapminder %>%
  filter(year == 1997) %>%
  ggplot(aes(x = pop, y = lifeExp)) +
  geom_point(shape = 21, color = 'red') +
  geom_smooth(method = 'loess', se = F, color = 'black') +
  labs(x = 'Population',
       y = 'Life expectancy, years',
       title = 'Life expectancy by population',
       caption = 'Data: Gapminder, 1997') +
  scale_x_log10(labels = scales::label_log()) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())

```

---
# Bias due to reverse causality

- Exists when the outcome is acting on the proposed cause. $X \leftrightarrow Y$
- If $Y$ determines exposure to $X$, then the correlation reflects the effect of $Y$ on $X$ in full or in part.

> Could life expectancy be acting on GDP?


---
# Bias due to confounding


- Selection/confounding bias exists when:
  - Treatment groups differ systematically by some factor, $Z$
  - *AND* $Z$ is a cause of $Y$
- In this case, the estimated effect of $X$ on $Y$ erroneously captures the influence of $Z$.

> Is there a variable related to GDP that might also, independently, impact life expectancy?

